---
title: "SWIRE_CAPSTONE"
author: "Imogen Holdsworth"
date: "2025-01-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

pacman::p_load(pacman,dplyr,caret, tidyverse,ggplot2, readxl,maps,tidyr,readr)

```

## Customer Address and Zip Mapping:

```{r}
# Load the data
AddressZip_OG <- read.csv("/Users/u0847758/Desktop/CAP/customer_address_and_zip_mapping.csv")
```


```{r}
# Ensure AddressZip_OG is a data frame with the correct column structure
# Assume the relevant column to split is named 'full_address'

# Split the column
AddressZip_Split <- AddressZip_OG |>
  separate(full.address, into = c("ZIP", "City", "State Name", "State Short", 
                                  "County","Code", "Latitude", "Longitude"), sep = ",")

# View the result
print(AddressZip_Split)

AddressZip_Split[rowSums(is.na(AddressZip_Split)) > 0, ]

# none missing

```


## Customer Profiles

```{r}
customer_profile<- read.csv("/Users/u0847758/Desktop/CAP/customer_profile.csv")
```


## Delivery Cost Data

```{r}
delivery_cost_data <- read_excel("/Users/u0847758/Desktop/CAP/delivery_cost_data (1).xlsx")
```

## Transactional 
```{r}
transactional_data<-  read.csv("/Users/u0847758/Desktop/CAP/transactional_data (1).csv")
```

```{r}

# Transactional Data EDA
colSums(is.na(transactional_data))
  # no missing

summary(transactional_data)

#ordered cases has a super high 8479.89 max compared to a median7.00 of and a mean of26.85, ,loaded cases and delivered cases is similar. delivered cases has an odd negative min, same with delivered gallons

ggplot(transactional_data, aes(y = ORDERED_CASES)) +
  geom_boxplot() +
  labs(title = "Boxplot of Ordered Cases", y = "Ordered Cases")

ggplot(transactional_data, aes(x = ORDERED_CASES)) +
  geom_density(fill = "purple", alpha = 0.5) +
  theme_minimal() +
  labs(title = "Density Plot of Ordered Cases", x = "Ordered Cases")

transactional_data |>
  arrange(desc(ORDERED_CASES)) |>
  head(50)
# 8479 isnt crazy i guess, may need to do some sort of transformations maybe log to normalize. but the third quartile is only 18.50 

```



```{r}
colSums(is.na(customer_profile))

  #PRIMARY_GROUP_NUMBER  missing 18196

summary(customer_profile)


# trade and cold drink cahnnel sub trade channel should maybe be factors, CO2 customer should be binary indicator, same with local market partner
```

```{r}
colSums(is.na(delivery_cost_data))
summary(delivery_cost_data)

#Cold Drink Channel, Vol Range, Cost Type <- should maybe be factors (ordinal or categorial maybe?)


```

```{r}
# IQR for outliers:

Q1 <- quantile(transactional_data$ORDERED_CASES, 0.25, na.rm = TRUE)
Q3 <- quantile(transactional_data$ORDERED_CASES, 0.75, na.rm = TRUE)
IQR_value <- Q3 - Q1

# Identify outliers
outliers <- transactional_data|>
  filter(ORDERED_CASES < (Q1 - 1.5 * IQR_value) | ORDERED_CASES > (Q3 + 1.5 * IQR_value))

print(nrow(outliers))  # Count of outliers
head(outliers)  # View some outliers
```
Summary Table:
 
collapsing transaction data into one row per customer per year

```{r}
# New summary table

customer_yearly_order_totals <- transactional_data |>
  group_by(YEAR, CUSTOMER_NUMBER) |>
  summarise(
    ORDERED_CASES = sum(ORDERED_CASES),
    DELIVERED_CASES = sum(DELIVERED_CASES),
    ORDERED_GALLONS = sum(ORDERED_GALLONS),
    DELIVERED_GALLONS = sum(DELIVERED_GALLONS)
  ) |>
  
# indicator binary for if the customer has reached or overaged on the threshold set by Swire
  
  mutate(OVER_400_GALLONS = ifelse(ORDERED_GALLONS > 400, 1, 0))  # Binary indicator

# View the first few rows
head(customer_yearly_order_totals)
```

compare the groups where customers reached or over threshold vs those who did not

```{r}
customer_yearly_order_totals |>
  group_by(OVER_400_GALLONS) |>
  summarise(
    AVG_ORDERED_CASES = mean(ORDERED_CASES),
    AVG_DELIVERED_CASES = mean(DELIVERED_CASES),
    AVG_ORDERED_GALLONS = mean(ORDERED_GALLONS),
    AVG_DELIVERED_GALLONS = mean(DELIVERED_GALLONS),
    COUNT_CUSTOMERS = n()  # Total customers in each group
  )


```

median instead of mean?

```{r}
customer_yearly_order_totals |>
  group_by(OVER_400_GALLONS) |>
summarise(
  MEDIAN_ORDERED_CASES = median(ORDERED_CASES)
)

ggplot(customer_yearly_order_totals, aes(x = as.factor(OVER_400_GALLONS), y = ORDERED_GALLONS)) +
  geom_boxplot(fill = "blue") +
  labs(title = "Distribution of Ordered Gallons by Customer Group")



```
^ this visualization is kind of worthless, but i cant decide what thresgold we shoudl set in regards to the outliers, 


```{r}

customer_yearly_order_totals|>
  group_by(OVER_400_GALLONS) |>
  summarise(AVG_ORDERED_GALLONS = mean(ORDERED_GALLONS, na.rm = TRUE)) |>
  ggplot(aes(x = as.factor(OVER_400_GALLONS), y = AVG_ORDERED_GALLONS, fill = as.factor(OVER_400_GALLONS))) +
  geom_col() +
  labs(title = "Average Ordered Gallons by Customer Group", 
       x = "Over 400 Gallons (0 = No, 1 = Yes)", 
       y = "Average Ordered Gallons") +
  scale_fill_manual(values = c("red", "blue"), name = "Over 400 Gallons") +
  theme_minimal()


```
ordered vs delivers? 
